---
title: "Metabolic-Calculations-Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metabolic-Calculations-Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tabletools)

```

# Introduction

This package has evolved to include functions useful for metabolic studies in humans, in particular to assess insulin sensitivity.

# Unit conversion functions

Convenience functions are provided to convert glucose, insulin and other units to the desired units.

## Glucose and Insulin

Most functions default to using glucose (mg/dL) and insulin (uU/mL) as the input units because these are most commonly reported in the US.

To convert glucose from mg/dL to mmol/L, use the `convert_glucose_to_mM` function, indicating the original units of the glucose value.

```{r}
convert_glucose_to_mM(100, glucose_units =  "mg/dL")
```

To convert insulin from uU/mL to pmol/L, use the `convert_insulin_to_pM` function, indicating the original units of the insulin value.

```{r}
convert_insulin_to_pM(10, insulin_units = "uU/mL")
```

## Weight and Height

Similar functions are available to convert weight and height to the desired units.

```{r}
convert_weight_to_kg(150, weight_units = "lbs")
```

Height from inches to meters:

```{r}
convert_height_to_m(70, height_units = "in")
```

# Insulin sensitivity

Sample OGTT Data, to use for fasting and OGTT based calculations.

```{r}
# handling data stored in a dataframe
ogtt1 <- data.frame(time=c(0, 30, 60, 90, 120),              # minutes
                    glucose=c(93, 129, 178, 164, 97),        # mg/dL
                    insulin=c(12.8, 30.7, 68.5, 74.1, 44.0)) # uU/mL

# example from Gutch et al 2015
ogtt2 <- data.frame(time=c(0, 30, 60, 90, 120),              # minutes
                    glucose=c(100, 160, 160, 160, 140),      # mg/dL
                    insulin=c(5, 10, 10, 10, 5))             # uU/mL
```

## Fasting Glucose and insulin

The HOMA-IR and QUICKI indices are used to assess insulin sensitivity. There is an updated HOMA2 calculation that is preferable to the original HOMA-IR calculation, but it is based on a series of equations that are not accessible to incorporate into a function. There is an online and downloadable calculator for HOMA2 [that is readily available here](https://www.rdm.ox.ac.uk/about/our-clinical-facilities-and-units/DTU/software/homa).

### A1C estimated average glucose

The estimated average glucose can be calculated from A1C:

```{r}
calculate_avg_glucose(5.5)
```

### HOMA-IR:

Higher values indicate lower insulin sensitivity.

```{r}
calculate_homair(ogtt1$glucose[1], ogtt1$insulin[1])
```

### QUICKI:

Higher values indicate higher insulin sensitivity.

```{r}
calculate_quicki(ogtt1$glucose[1], ogtt1$insulin[1])
```

## OGTT based calculations

### Matsuda Index

Matsuda index is the most often used OGTT-based estimate of insulin sensitivity. Calculations can be verified at an online calculator [here](https://mmatsuda.diabetes-smc.jp/MIndex.html).

```{r}
calculate_matsuda_index(ogtt1$time, ogtt1$glucose, ogtt1$insulin)
```

### Stumvoll Index

```{r}
calculate_stumvoll_isi(ogtt1$time, ogtt1$glucose, ogtt1$insulin) #
```

# Body Surface Area

The `bsa` command calculates Body Surface area (m<sup>2</sup>) using the most commonly available methods [Yu et al](https://pubmed.ncbi.nlm.nih.gov/19900761/).

The default method is the Mosteller formula. Functions will conduct unit conversions if the proper units are provided- defaults to kg and m.

```{r}
calculate_bsa(weight = 70, height =  1.50, method = "Mosteller",
              weight_units = "kg", height_units = "m")
```

Available Methods and Formulas:

```{=tex}
\begin{equation}
\sqrt{\frac{Ht(cm) \cdot Wt(kg)}{3600}} \tag{Mosteller}
\end{equation}
```
```{=tex}
\begin{equation}
Wt(kg)^{0.425} \cdot Ht(cm)^{0.725} \cdot 0.007184 \tag{DuBois and DuBois}
\end{equation}
```
```{=tex}
\begin{equation}
Wt(kg)^{0.51456} \cdot Ht(cm)^{0.42246} \cdot 0.0235 \tag{Gehan and George}
\end{equation}
```
```{=tex}
\begin{equation}
Wt(kg)^{0.5378} \cdot Ht(cm)^{0.3964} \cdot 0.024265 \tag{Haycock}
\end{equation}
```
```{=tex}
\begin{equation}
71.3989 \cdot Ht(cm)^{0.7437} \cdot Wt(kg)^{0.4040} \div 10000 \tag{Yu}
\end{equation}
```
```{=tex}
\begin{equation}
0.1173 \cdot Wt(kg)^{0.6466} \tag{Livingston}
\end{equation}
```
```{=tex}
\begin{equation}
128.1 \cdot Ht(cm)^{0.60} \cdot Wt(kg)^{0.44} \tag{Tikuisis}
\end{equation}
```
# Kidney function equations

## eGFR

Creatinine clearance is the historical standard to estimate glomerular filtration rate (GFR). The Cockcroft-Gault formula was used in the past, but has been superceded by the MDRD and more recently CKD-EPI equations. Newer methods incorporating Cystatin C are also available, and may increase accuracy when using race-free equations [@Inker2021].

Formulas for multiple methods of eGFR are provided- not meant to be comprehensive, but up to date. The ASN and NKF task force panels now recommend using equations that do not incorporate a race coefficient. Details on the formulas can be found in Supplemental Table S10 of [@Inker2021].

Online calculators to verify the results are available:

-   New CKD-EPI equations at [Kidney.org](https://www.kidney.org/professionals/kdoqi/gfr_calculator)\
-   Old 2009 CKD-EPI (race-based) calculator is maintained at [NIH/NIDDK](https://www.niddk.nih.gov/health-information/professionals/clinical-tools-patient-management/kidney-disease/laboratory-evaluation/glomerular-filtration-rate-calculators/historical) for comparison.
-   MDRD

### Cockcroft-Gault

```{r}
calculate_egfr_cg(age=70, sex="Male", weight=70, creatinine=1.0) # 95.2
```

### CKD-EPI

Note: When comparing results using this function against Inker Table S11 there are minor differences amounting to no more than 0.5-1.0 ml/min/1.73m2.

The [2021 CKD-EPI equation](https://www.kidney.org/content/ckd-epi-creatinine-equation-2021) is: 
$$eGFR_{cr}= 142\cdot min(Cr/\kappa, 1)^\alpha \cdot max(Cr/\kappa, 1)^{-1.200} \cdot 0.9938^{Age} \cdot 1.102[Female] $$ 

where:

\* $\kappa$ = 0.7 (Female) or 0.9 (Male)\
\* $\alpha$ = -0.241 (Female) or -0.302 (Male)\
\* Age is in years \* Cr = serum Creatinine in mg/dL

Note that the 2021 CKD-EPI formula also has a version which was derived using Race, and produces slightly different results. 

$$eGFR_{cr}= 141\cdot min(Cr/\kappa, 1)^\alpha \cdot max(Cr/\kappa, 1)^{-1.209} \cdot 0.9929^{Age} \cdot 1.018[Female] \cdot 1.159[Black] $$ 

where:

\* $\kappa$ = 0.7 (Female) or 0.9 (Male)\
\* $\alpha$ = -0.329 (Female) or -0.411 (Male)\

```{r}
calculate_egfr_ckdepi(age=70, sex="Male", creatinine=0.8) # 95.2
```



### CKD-EPI Cr-Cystatin-C

not yet incorporated
