---
title: "Metabolic-Calculations-Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metabolic-Calculations-Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tabletools)

```

# Introduction

This package has evolved to include functions useful for metabolic studies in humans, in particular to assess insulin sensitivity. 

# Unit conversion functions

Convenience functions are provided to convert glucose, insulin and other units to the desired units. 

## Glucose and Insulin

Most functions default to using glucose (mg/dL) and insulin (uU/mL) as the input units because these are most commonly reported in the US.

To convert glucose from mg/dL to mmol/L, use the `convert_glucose_to_mM` function, indicating the original units of the glucose value. 

```{r}
convert_glucose_to_mM(100, glucose_units =  "mg/dL")
```

To convert insulin from uU/mL to pmol/L, use the `convert_insulin_to_pM` function, indicating the original units of the insulin value. 

```{r}
convert_insulin_to_pM(10, insulin_units = "uU/mL")
```

## Weight and Height

Similar functions are available to convert weight and height to the desired units.
```{r}
convert_weight_to_kg(150, weight_units = "lbs")
```

Height from inches to meters:
```{r}
convert_height_to_m(70, height_units = "in")
```



# Insulin sensitivity

Sample OGTT Data, to use for fasting and OGTT based calculations. 

```{r}
# handling data stored in a dataframe
ogtt1 <- data.frame(time=c(0, 30, 60, 90, 120),              # minutes
                    glucose=c(93, 129, 178, 164, 97),        # mg/dL
                    insulin=c(12.8, 30.7, 68.5, 74.1, 44.0)) # uU/mL

# example from Gutch et al 2015
ogtt2 <- data.frame(time=c(0, 30, 60, 90, 120),              # minutes
                    glucose=c(100, 160, 160, 160, 140),      # mg/dL
                    insulin=c(5, 10, 10, 10, 5))             # uU/mL
```



## Fasting Glucose and insulin

The HOMA-IR and QUICKI indices are used to assess insulin sensitivity.
There is an updated HOMA2 calculation that is preferable to the original HOMA-IR calculation, but it is based on a series of equations that are not accessible to incorporate into a function. There is an online and downloadable calculator for HOMA2 [that is readily available here](https://www.rdm.ox.ac.uk/about/our-clinical-facilities-and-units/DTU/software/homa).

### A1C estimated average glucose

The estimated average glucose can be calculated from A1C:
```{r}
calculate_avg_glucose(5.5)
```


### HOMA-IR: 

Higher values indicate lower insulin sensitivity. 

```{r}
calculate_homair(ogtt1$glucose[1], ogtt1$insulin[1])
```


### QUICKI:

Higher values indicate higher insulin sensitivity. 


```{r}
calculate_quicki(ogtt1$glucose[1], ogtt1$insulin[1])
```

## OGTT based calculations

### Matsuda Index

Matsuda index is the most often used OGTT-based estimate of insulin sensitivity. 
Calculations can be verified at an online calculator [here](https://mmatsuda.diabetes-smc.jp/MIndex.html).


```{r}
calculate_matsuda_index(ogtt1$time, ogtt1$glucose, ogtt1$insulin)
```

### Stumvoll Index

```{r}
calculate_stumvoll_isi(ogtt1$time, ogtt1$glucose, ogtt1$insulin) #
```

# Body Surface Area  

The `bsa` command calculates Body Surface area (m<sup>2</sup>) using the most commonly available methods [Yu et al](https://pubmed.ncbi.nlm.nih.gov/19900761/). 

The default method is the Mosteller formula. 
Functions will conduct unit conversions if the proper units are provided- defaults to kg and m.
```{r}
calculate_bsa(weight = 70, height =  1.50, method = "Mosteller",
              weight_units = "kg", height_units = "m")
```


Available Methods and Formulas:  

\begin{equation}
\sqrt{\frac{Ht(cm) \cdot Wt(kg)}{3600}} \tag{Mosteller}
\end{equation}

\begin{equation}
Wt(kg)^{0.425} \cdot Ht(cm)^{0.725} \cdot 0.007184 \tag{DuBois and DuBois}
\end{equation}

\begin{equation}
Wt(kg)^{0.51456} \cdot Ht(cm)^{0.42246} \cdot 0.0235 \tag{Gehan and George}
\end{equation}

\begin{equation}
Wt(kg)^{0.5378} \cdot Ht(cm)^{0.3964} \cdot 0.024265 \tag{Haycock}
\end{equation}

\begin{equation}
71.3989 \cdot Ht(cm)^{0.7437} \cdot Wt(kg)^{0.4040} \div 10000 \tag{Yu}
\end{equation}

\begin{equation}
0.1173 \cdot Wt(kg)^{0.6466} \tag{Livingston}
\end{equation}

\begin{equation}
128.1 \cdot Ht(cm)^{0.60} \cdot Wt(kg)^{0.44} \tag{Tikuisis}
\end{equation}

