---
title: "Metabolic-Calculations-Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metabolic-Calculations-Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tabletools)

```

# Body Surface Area  

The `bsa` command calculates Body Surface area (m<sup>2</sup>) using the most commonly available methods [Yu et al](https://pubmed.ncbi.nlm.nih.gov/19900761/). 

Available Methods and Formulas:  

\begin{equation}
\sqrt{\frac{Ht(cm) \cdot Wt(kg)}{3600}} \tag{Mosteller}
\end{equation}

\begin{equation}
Wt(kg)^{0.425} \cdot Ht(cm)^{0.725} \cdot 0.007184 \tag{DuBois and DuBois}
\end{equation}

\begin{equation}
Wt(kg)^{0.51456} \cdot Ht(cm)^{0.42246} \cdot 0.0235 \tag{Gehan and George}
\end{equation}

\begin{equation}
Wt(kg)^{0.5378} \cdot Ht(cm)^{0.3964} \cdot 0.024265 \tag{Haycock}
\end{equation}

\begin{equation}
71.3989 \cdot Ht(cm)^{0.7437} \cdot Wt(kg)^{0.4040} \div 10000 \tag{Yu}
\end{equation}

\begin{equation}
0.1173 \cdot Wt(kg)^{0.6466} \tag{Livingston}
\end{equation}

\begin{equation}
128.1 \cdot Ht(cm)^{0.60} \cdot Wt(kg)^{0.44} \tag{Tikuisis}
\end{equation}

