---
title: "Simple examples"
author: "JM Luther"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
library(dplyr)
library(ggplot2)
library(htmlTable)
library(tabletools)
library(tidyr)
```

# tabletools

The goal of tabletools is to provide functions which are helpful for generating summary html tables.
For details on the functions, please see help documentation in the package.

## Installation

Install from Github:  `devtools::install_github("JMLuther/tabletools")`


These functions are most useful in conjunction with an html table generating package::functions such as:  
  
  *  `htmlTable::htmlTable` [Tables with htmlTable and some alternatives](https://cran.r-project.org/web/packages/htmlTable/vignettes/tables.html)  
  *  `knitr::kable` see [Create Awesome HTML Table with knitr::kable and kableExtra](https://haozhu233.github.io/kableExtra/awesome_table_in_html.html)  
  *  `formattable` 

## `my_summary` custom summary  

This summary function is customized to my preferences, and should provide enough data to pipe into ggplot or a simple summary table.
Just provides a summary for a single continuous variable.

Table:
```{r}
mtcars %>% 
  group_by(cyl) %>% 
  my_summary(mpg, digits = 1) %>% # override the default 2 decimal point rounding
  htmlTable::htmlTable(rnames=F, align = "c",
                       css.cell = rbind(rep("padding-left: .5em; padding-right: .5em;",times=ncol(.)),
                                        matrix("padding:0 5px 0 5px;", ncol=ncol(.), nrow=nrow(.))))

```

Plot:
```{r}
mtcars %>% 
  group_by(cyl) %>% 
  my_summary(mpg) %>% 
  ggplot(aes(factor(cyl), mpg_mean,
             ymin = mpg_ci_025, ymax = mpg_ci_975)) +
  geom_linerange() +
  geom_point(color = "red", size=2) +
  labs(x="No. of Cylinders",
       y= "Average MPG",
       title= "Average MPG by number of Cylinders") +
  theme_bw()

```


# Summary text functions 

```{r}
mtcars %>%
  summarise(mean_sd = txt_mean_sd(mpg),
            mean_sem = txt_mean_sem(mpg),
            mean_range = txt_mean_range(mpg),
            median_iqr = txt_median_iqr(mpg),
            pct_fr = txt_pct_fr(vs, 1)
            ) %>% 
  tidyr::gather(txt_f, output) %>% # this is better than using t() which creates a matrix
  htmlTable::htmlTable(rnames=F, align = "l", align.header = "l", 
                       caption = "Example results",
                       css.cell = rbind(rep("padding-left: .5em; padding-right: .5em;",times=ncol(.)),
                                        matrix("padding:0 5px 0 5px;", ncol=ncol(.), nrow=nrow(.))))
  

```




The following functions provide useful
# Session Info  

<button class="btn btn-primary" data-toggle="collapse" data-target="#SessionInfo"> Show/Hide </button>  
<div id="SessionInfo" class="collapse">  
```{r echo=F}
Hmisc::markupSpecs$html$session()
```
</div>
 